First Order Alternation

### 2018-10-22 quantifiers/draft.tex
ok

\begin{abstract}
\end{abstract}

@@

Alternating automata are a generalization of nondeterministic automata
with universal transitions, that create several copies of the
automaton, which synchronize on the same input word. Alternating
automata are appealing for verification because they allow encoding of
problems such as temporal logic model checking in linear time, as
opposed to the exponential time required by nondeterministic automata
\cite{Vardi95}.

%% However, checking the emptiness of an alternating automaton over a
%% finite alphabet is \pspace-complete, instead of
%% \nlogspace-complete, which is the case for nondeterministic automata.
%% In practice, this complexity can be dealt with using efficient
%% antichain algorithms \cite{Ganty09}.

A finite-alphabet alternating automaton is typically described by
transitions $q \arrow{a}{} \phi$, where $q$ is a state, $a$ is an
input symbol and $\phi$ is a positive boolean combinations of states,
viewed as propositional variables. Here we introduce a generalized
alternating automata model in which states are predicate symbols
$q(y_1,\ldots,y_k)$, the input has associated data variables
$x_1,\ldots,x_n$, ranging over an infinite domain and transitions are
of the form $q(y_1,\ldots,y_k) \arrow{a(x_1,\ldots,x_n)}{} \phi$,
where $\phi$ is any formula in the first-order theory of the data
domain, in which each state predicate occurs under an even number of
negations. In this model, the arguments of a predicate atom
$q(y_1,\ldots,y_k)$ track the values of the \emph{internal variables}
associated with the state. Together with the input values
$x_1,\ldots,x_n$, these values are used to compute the successor
states and are invisible in the input sequence.

Previous attempts to generalized classical Rabin-Scott automata to
infinite alphabets, such as timed automata \cite{AlurDill94} and
finite-memory (register) automata \cite{KaminskiFrancez94} face the
\emph{complement closure} problem: there exist automata for which the
complement language cannot be recognized by an automaton in the same
class. This excludes the possibility of encoding a language inclusion
problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
denotes the complement of $\lang{B}$. However, when all internal
variables are made visible in the input, determinization and
complementation are possible, for data theories closed under negation,
such as integer or real arithmetics \cite{IosifRV16}.

The solution we adopt is a tight coupling of internal variables to
control states. As we show, this allows for linear-time
complementation just as in the case of boolean alternating
automata. Complementation is, moreover, possible when the transition
formulae contain first-order quantifiers, generating
infinitely-branching executions.

The price to be paid for this expressivity is that emptiness of
first-order alternating automata is undecidable, even for the simplest
data theory of equality \cite{Farzan15}. We work around this problem
by developing an abstraction refinement lazy annotation semi-algorithm
in the spirit of the \impact~procedure for checking safety of
nondeterministic integer programs \cite{McMillan06}. However, checking
emptiness of first-order alternating automata by lazy annotation with
interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules are problematic to decide if a given
  trace is spurious, mainly because adding uninterpreted predicate
  symbols to decidable first-order theories, such as Presburger
  arithmetic, results in undecidability \cite{Halpern91}. We deal with
  this problem by assuming that the first-order data theory, without
  uninterpreted predicate symbols, has an effective quantifier
  elimination procedure, that instantiates quantifers with witness
  terms.
%
\item The interpolants used to prove the spuriousness of an input
  sequence are not \emph{local} anymore, as they may refer to input
  values encountered in the past. However, the future executions are
  oblivious to when these values have been seen in the past and depend
  only on the data constraints between the values. We use this
  observation to define a labeling of nodes visited by the lazy
  annotation semi-algorithm as conjunctions of existentially
  quantified interpolants that combine predicate symbols with data
  constraints.
\end{compactenum}

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata. Further, we
describe applications of this method to several undecidable but
recursively enumerable problems, for which no effective
semi-algorithms exist, to date: inclusion between recognizable timed
languages \cite{AlurDill94} and between languages recognized by
finite-memory automata \cite{KaminskiFrancez94}, as well as emptiness
of predicate automata, a model of first-order alternating automata
used to check safety and liveness properties of parameterized
concurrent programs \cite{Farzan15,Farzan16}.

\paragraph{\bf Related Work} 
The first-order alternating automata model presented in this paper
stems from our previous work on boolean alternating automata extended
with variables ranging over infinite data \cite{IosifXu18}. There we
considered that states are propositional variables, as in the
classical textbook alternating automata model, and that all the
variables of the automaton are reflected in the input. The model in
this paper overcomes this latter restriction by allowing for internal
variables, whose variables are not visible in the language. This
solves an older language inclusion problem $\bigcap_{i=1}^n\lang{A_i}
\subseteq \lang{B}$, between finite-state automata with data variables
\cite{IosifRV16}. There, we assumed that all variables of the observer
automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
model the concurrent components of the system under check.

@@

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first-order
formulae is defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. The size
$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
write it down.

@@

A formula is in \emph{prenex form} if it is of the form $\varphi =
Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
first-order formula can be written in prenex form, by renaming each
quantified variable to a unique name and moving the quantifiers
upfront.

@@

Formally, a \emph{first-order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
occurring in the right-hand side formula of a transition rule are
referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first-order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$.

@@

To tackle this problem, we start from the observation that acceptance
formulae have a particular form, which allows the elimination of path
quantifiers and of predicates, by a couple of
satisfiability-preserving transformations. The result of applying
these transformations is a formula with no predicate symbols, whose
only quantifiers are those introduced by the transition rules of the
automaton, referred to as transition quantifiers. We shall further
assume (\S\ref{sec:emptiness}) that the first-order theory of the data
sort $\Data$ has quantifier elimination, which allows to effectively
decide the satisfiability of such formulae.

@@

We check emptiness of first-order alternating automata using a version
of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
analogous procedure is given in \cite{IosifXu18}, for a simpler model
of alternating automata, that uses only predicates or arity zero
(booleans) and no transition quantifiers. For simplicity, we do not
present the details of this algorithm (see Appendix
\ref{app:pseudocode}) and shall content ourselves of several
high-level definitions.

@@

Lazy annotation semi-algorithms \cite{McMillan06,McMillan14} build
unfoldings of automata trying to discover counterexamples for
emptiness. If the automaton $\A$ in question is non-empty, a
systematic enumeration of the input event sequences\footnote{For
  instance, using breadth-first search.} from $\Sigma^*$ will suffice
to discover a word $w \in \lang{\A}$, provided that the first-order
theory of the data domain $\Data$ is decidable (Lemma
\ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$, the
enumeration of input event sequences may, in principle, run
forever. The typical way of fighting this divergence problem is to
define a \emph{coverage} relation between the nodes of the unfolding
tree.

@@

As mentioned above, we check emptiness of first-order alternating
automata using the same method previously used to check emptiness of a
simpler model of alternating automata, which uses boolean constants
for control states and whose transition rules have no quantifiers
\cite{IosifXu18}. The higher complexity of the automata model
considered here, manifests itself within the interpolant generation
procedure, used to refine the labeling of the unfolding. We discuss
generation of interpolants in the next section.

@@

\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
\pset{-}{\phi} \cap \pset{-}{\psi}$.
\end{definition}
%% The existence of Lyndon interpolants for first-order logic, with no
%% predefined interpretation of function symbols\footnote{A model
%%   specifies the universe and the interpretation of function symbols.},
%% was proved by a non-constructive model-theoretic argument
%% \cite{Lyndon59}. In the following, we shall consider only those
%% theories of the data domain $\Data$, for which there exists a decision
%% procedure for satisfiability, able to produce Lyndon interpolants. For
%% instance, such interpolating decision procedures exist for linear real
%% and integer arithmetic, with uninterpreted function symbols (UFLRA and
%% UFLIA).

@@

This is where the predicate-free path formulae (Definition
\ref{def:substform}) come into play. For a given event sequence
$\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
\alpha$ if and only if $\substform{\alpha}$ is satisfiable. Assuming
further that the equality atoms in the transition rules of $\A$ are
written in the language of a decidable first-order theory, such as
Presburger arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives
us an effective way of checking emptiness of $\A$, relative to a given
event sequence. However, this method does not cope well with lazy
annotation, because there is no way to extract, from the
unsatisfiability proof of $\substform{\alpha}$, the interpolants
needed to annotate $\alpha$. This is because \begin{inparaenum}[(i)]
\item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
  repeated substitutions (Definition \ref{def:substform}) loses track
  of the steps of the execution, and
%
\item\label{it2:gli} quantifiers that occur nested in
  $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
  as an unsatisfiable conjunction of formulae from which interpolants
  are extracted (Definition \ref{def:lyndon-interpolant}).
\end{inparaenum}

@@

Under the assumption that witness terms can be effectively built, let
us describe the generation of a non-local GLI for a given input event
sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
the acceptance formula $\accform{\alpha}$ and its equisatisfiable
forms $\quantform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$
and $\substform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$,
both written in prenex form, with matrices $\widehat{\Phi}$ and
$\overline{\Phi}$, respectively. Because we assumed that the
first-order theory of $\Data$ has quantifier elimination, the
satisfiability problem for $\substform{\alpha}$ is decidable. If
$\substform{\alpha}$ is satisfiable, we build a counterexample for
emptiness $w$ such that $\event{w}=\alpha$ and $\data{w}$ is a
satisfying assignment for $\substform{\alpha}$. Otherwise,
$\substform{\alpha}$ is unsatisfiable and there exist witness terms
$\tau_{i_1} \ldots \tau_{i_\ell}$, where $\set{i_1, \ldots, i_\ell} =
\set{j \in [1,m] \mid Q_j = \forall}$, such that
$\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
\ref{thm:skolem}). Then it turns out that the formula
$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
$\quantform{\alpha}$, is unsatisfiable as well (Lemma
\ref{lemma:subst-quant}). Because this latter formula is structured as
a conjunction of formulae $\stamp{\iota}{0} \wedge \phi_1 \ldots
\wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap \stamp{Q}{\leq n}
\subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and $\voc{\psi} \cap
\stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now possible to use
an existing interpolation procedure for the quantifier-free theory of
$\Data$, extended with uninterpreted function symbols, to compute a
sequence of non-local GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
\cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.

@@

  we assumed that the first-order theory of the data domain has

@@

In conclusion, under two assumptions about the first-order theory of
the data domain, namely the \begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
to use such non-local interpolants to label the unfolding of an
automaton (Definition \ref{def:unfolding}) and to compute the coverage
between nodes of the unfolding (Definition \ref{def:coverage}).

@@

The main application of first-order alternating automata is checking
inclusion between various classes of automata extended with variables
ranging over infinite domains that recognize languages over infinite
alphabets. The most widely known such classes are \emph{timed
  automata} \cite{AlurDill94} and \emph{finite-memory (register)
  automata} \cite{KaminskiFrancez94}. In both cases, complementation
is not possible inside the class and inclusion is undecidable. Our
contribution is providing a systematic semi-algorithm for these
decision problems. In addition, the method described in
\S\ref{sec:emptiness} can extend our previous \emph{generic register
  automata} \cite{IosifRV16} inclusion checking framework, by allowing
monitor (right-hand side) automata to have local variables, that are
not visible in the language.

>>>>

\begin{abstract}
We introduce first order alternating automata, a generalization of
boolean alternating automata, in which transition rules are described
by multisorted first order formulae, with states and internal
variables given by uninterpreted predicate terms. The model is closed
under union, intersection and complement, and its emptiness problem is
undecidable, even for the simplest data theory of equality. To cope
with this limitation, we develop an abstraction refinement
semi-algorithm based on lazy annotation of the symbolic execution
paths with interpolants, obtained by applying (i) quantifier
elimination with witness term generation and (ii) Lyndon interpolation
in the quantifier-free data theory with uninterpreted predicate
symbols. This provides a method for checking inclusion of timed and
finite-memory register automata, and emptiness of quantified predicate
automata, previously used in the verification of parameterized
concurrent programs, composed of replicated threads, with a
shared-memory communication model.
\end{abstract}

@@

Alternating automata are a generalization of nondeterministic automata
with universal transitions, that create several copies of the
automaton, which synchronize on the same input word. Alternating
automata are appealing for verification because they allow encoding of
problems such as temporal logic model checking in linear time, as
opposed to the exponential time required by nondeterministic automata
\cite{Vardi95}. A finite-alphabet alternating automaton is typically
described by a set of transition rules $q \arrow{a}{} \phi$, where $q$
is a state, $a$ is an input symbol and $\phi$ is a positive boolean
combinations of states, viewed as propositional variables.

Here we introduce a generalized alternating automata model in which
states are predicate symbols $q(y_1,\ldots,y_k)$, the input has
associated data variables $x_1,\ldots,x_n$, ranging over an infinite
domain and transitions are of the form $q(y_1,\ldots,y_k)
\arrow{a(x_1,\ldots,x_n)}{} \phi$, where $\phi$ is any formula in the
first-order theory of the data domain, in which each state predicate
occurs under an even number of negations. In this model, the arguments
of a predicate atom $q(y_1,\ldots,y_k)$ track the values of the
\emph{internal variables} associated with the state. Together with the
input values $x_1,\ldots,x_n$, these values are used to compute the
successor states and are invisible in the input sequence.

Previous attempts to generalize classical Rabin-Scott automata to
infinite alphabets, such as timed automata \cite{AlurDill94} and
finite-memory (register) automata \cite{KaminskiFrancez94} face the
\emph{complement closure} problem: there exist automata for which the
complement language cannot be recognized by an automaton in the same
class. This excludes the possibility of encoding a language inclusion
problem $\lang{A} \subseteq \lang{B}$ as the emptiness of an automaton
recognizing the language $\lang{A} \cap \clang{B}$, where $\clang{B}$
denotes the complement of $\lang{B}$.

The solution we adopt here is a tight coupling of internal variables
to control states, using uninterpreted predicate symbols. As we show,
this allows for linear-time complementation just as in the case of
boolean alternating automata. Complementation is, moreover, possible
when the transition formulae contain first-order quantifiers,
generating infinitely-branching execution trees. The price to be paid
for this expressivity is that emptiness of first-order alternating
automata is undecidable, even for the simplest data theory of equality
\cite{Farzan15}.

The main contribution of this paper is an effective emptiness checking
semi-algorithm for first-order alternating automata, in the spirit of
the \impact~procedure, originally developed for checking safety of
nondeterministic integer programs \cite{McMillan06}. Essentially,
checking emptiness of first-order alternating automata by lazy
annotation with interpolants faces two problems: \begin{compactenum}
\item Quantified transition rules make it hard, or even impossible, to
  decide if a given symbolic trace is spurious. This is mainly because
  adding uninterpreted predicate symbols to decidable first-order
  theories, such as Presburger arithmetic, results in undecidability
  \cite{Halpern91}. To deal with this problem, we assume that the
  first order data theory, without uninterpreted predicate symbols,
  has a quantifier elimination procedure, that instantiates quantifers
  with effectively computable \emph{witness terms}.
%
\item The interpolants that prove the spuriousness of a symbolic path
  are not \emph{local}, as they may refer to input values encountered
  in the past. However, the future executions are oblivious to
  \emph{when} these values have been seen in the past and depend only
  on the data constraints between the values. We use this fact to
  define a labeling of nodes, visited by the lazy annotation
  procedure, with conjunctions of existentially quantified
  interpolants combining predicate atoms with data constraints.
\end{compactenum}

We describe applications of this semi-algorithmic method to several
undecidable but recursively enumerable problems, for which no
effective method exist, to date: inclusion between recognizable timed
languages \cite{AlurDill94} and between languages recognized by
finite-memory automata \cite{KaminskiFrancez94}, as well as emptiness
of predicate automata, a model of first-order alternating automata
used to check safety and liveness properties of parameterized
concurrent programs \cite{Farzan15,Farzan16}.

\paragraph{\bf Related Work} 
The first order alternating automata model presented in this paper
stems from our previous work on boolean alternating automata extended
with variables ranging over infinite data \cite{IosifXu18}. There we
considered states to be propositional variables, as in the classical
textbook alternating automata model, and all variables of the
automaton to be observable in the input. The model in this paper
overcomes this latter restriction by allowing for internal variables,
whose variables are not visible in the language.

This solves an older language inclusion problem
$\bigcap_{i=1}^n\lang{A_i} \subseteq \lang{B}$, between finite-state
automata with data variables, whose languages are alternating
sequences of input events and variable valuations
\cite{IosifRV16}. There, we assumed that all variables of the observer
automaton $B$ must be declared in the automata $A_1, \ldots, A_n$ that
model the concurrent components of the system under check.

The work probably closest to the one reported here concerns the model
of predicate automata (PA) \cite{Farzan15,Farzan16,KincaidPhD},
applied to the verification of parameterized concurrent programs with
shared memory. In this model, the alphabet consists of pairs of
program statements and thread identifiers, thus being infinite because
the number of threads is unbounded. Because thread identifiers can
only be compared for (dis-)equality, the data theory in PA is the
theory of equality. Even with this simplification, the emptiness
problem is undecidable when either the predicates have arity greater
than one \cite{Farzan15} or quantified transition rules
\cite{KincaidPhD}. Checking emptiness of quantifier free PA is
possible semi-algorithmically, by explicitly enumerating reachable
configurations and checking coverage by looking for permuations of
argument values. However, no semi-algorithm is given for quantified
PA.  Dealing with quantified transition rules is one of the
contributions of the work reported in this paper.

@@

Let $\vars = \set{x,y,z,\ldots}$ be an infinite countable set of
variables, ranging over $\Data$. Terms are either constants of sort
$\Data$, variables or function applications $f(t_1,\ldots,t_{\#(f)})$,
where $t_1,\ldots,t_{\#(f)}$ are terms. The set of first order
formulae is defined by the syntax below:
\[\phi := t \teq s \mid p(t_1,\ldots,t_{\#(p)}) 
\mid \neg \phi_1 \mid \phi_1 \wedge \phi_2 \mid \exists x ~.~
\phi_1 \] where $t,s,t_1,\ldots,t_{\#(p)}$ denote terms. We write
$\phi_1 \vee \phi_2$, $\phi_1 \rightarrow \phi_2$ and $\forall x ~.~
\phi_1$ for $\neg(\neg\phi_1 \wedge \neg\phi_2)$, $\neg\phi_1 \vee
\phi_2$ and $\neg\exists x ~.~ \neg\phi_1$, respectively. We denote by
$\fv{}{\phi}$ the set of free variables in $\phi$. The size
$\len{\phi}$ of a formula $\phi$ is the number of symbols needed to
write it down.

@@

A formula is in \emph{prenex form} if it is of the form $\varphi =
Q_1x_1 \ldots Q_nx_n ~.~ \phi$, where $\phi$ has no quantifiers. In
this case we call $\phi$ the \emph{matrix} of $\varphi$. Every
first order formula can be written in prenex form, by renaming each
quantified variable to a unique name and moving the quantifiers
upfront.

@@

Formally, a \emph{first order alternating automaton} is a tuple $\A =
\tuple{\Sigma,X,Q,\iota,F,\Delta}$, where $\Sigma$ is a finite set of
input events, $X$ is a finite set of input variables, $Q$ is a finite
set of predicates denoting control states, $\iota \in
\posforms(Q,\emptyset)$ is a sentence defining initial configurations,
$F \subseteq Q$ is the set of predicates denoting final states, and
$\Delta$ is a set of \emph{transition rules} of the form
\(q(y_1,\ldots,y_{\#(q)}) \arrow{a(X)}{} \psi\), where $q \in Q$ is a
predicate, $a \in \Sigma$ is an input event and $\psi \in
\posforms(Q,X \cup \set{y_1,\ldots,y_{\#(q)}})$ is a positive formula,
where $X \cap \set{y_1,\ldots,y_{\#(q)}} = \emptyset$. The quantifiers
occurring in the right-hand side formula of a transition rule are
referred to as \emph{transition quantifiers}. The \emph{size} of $\A$
is defined as $\len{\A} = \len{\iota} +
\sum_{\scriptscriptstyle{q(\vec{y}) \arrow{a(X)}{} \psi \in \Delta}}
\len{\psi}$.

@@

Given a predicate $q \in Q$ and a tuple of data values
$d_1,\ldots,d_{\#(q)}$, the tuple $q(d_1,\ldots,d_{\#(q)})$ is called
a \emph{configuration}\footnote{Note that a configuration is not a
  logical term since data values cannot be written in logic.}. To
formalize the execution semantics of automata, we relate sets of
configurations to models of first order sentences, as follows. Given a
formula $\phi$ and a valuation of its free variables $\nu$, each
interpretation $\I \in \sem{\phi}_\nu$ corresponds to a set of
configurations $\cube{\I} \isdef \set{q(d_1,\ldots,d_{\#(q)}) \mid q
  \in Q,~ \tuple{d_1,\ldots,d_{\#(q)}} \in q^\I}$, called a
\emph{cube}. For a set $\mathcal{S}$ of interpretations, we define
$\cube{\mathcal{S}} \isdef \set{\cube{\I} \mid \I \in
  \mathcal{S}}$.

@@

To tackle this problem, we start from the observation that acceptance
formulae have a particular form, which allows the elimination of path
quantifiers and of predicates, by a couple of
satisfiability-preserving transformations. The result of applying
these transformations is a formula with no predicate symbols, whose
only quantifiers are those introduced by the transition rules of the
automaton, referred to as transition quantifiers. We shall further
assume (\S\ref{sec:emptiness}) that the first order theory of the data
sort $\Data$ has quantifier elimination, which allows to effectively
decide the satisfiability of such formulae.

@@

We check emptiness of first order alternating automata using a version
of the \impact~ lazy annotation semi-algorithm \cite{McMillan06}. An
analogous procedure is given in \cite{IosifXu18}, for a simpler model
of alternating automata, that uses only predicates or arity zero
(booleans) and no transition quantifiers. For simplicity, we do not
present the details of this algorithm (see Appendix
\ref{app:pseudocode}) and shall content ourselves of several
high-level definitions.

@@

Lazy annotation semi-algorithms \cite{McMillan06,McMillan14} build
unfoldings of automata trying to discover counterexamples for
emptiness. If the automaton $\A$ in question is non-empty, a
systematic enumeration of the input event sequences\footnote{For
  instance, using breadth-first search.} from $\Sigma^*$ will suffice
to discover a word $w \in \lang{\A}$, provided that the first order
theory of the data domain $\Data$ is decidable (Lemma
\ref{lemma:acceptance}). However, if $\lang{\A} = \emptyset$, the
enumeration of input event sequences may, in principle, run
forever. The typical way of fighting this divergence problem is to
define a \emph{coverage} relation between the nodes of the unfolding
tree.

@@

As mentioned above, we check emptiness of first order alternating
automata using the same method previously used to check emptiness of a
simpler model of alternating automata, which uses boolean constants
for control states and whose transition rules have no quantifiers
\cite{IosifXu18}. The higher complexity of the automata model
considered here, manifests itself within the interpolant generation
procedure, used to refine the labeling of the unfolding. We discuss
generation of interpolants in the next section.

@@

\begin{definition}[\cite{Lyndon59}]\label{def:lyndon-interpolant}
Given formulae $\phi$ and $\psi$ such that $\phi \wedge \psi$ is
unsatisfiable, a \emph{Lyndon interpolant} is a formula $I$ such that
$\phi \models I$, the formula $I \wedge \psi$ is unsatisfiable,
$\voc{I} \subseteq \voc{\phi} \cap \voc{\psi}$, $\pset{+}{I} \subseteq
\pset{+}{\phi} \cap \pset{+}{\psi}$ and $\pset{-}{I} \subseteq
\pset{-}{\phi} \cap \pset{-}{\psi}$.
\end{definition}
%% The existence of Lyndon interpolants for first order logic, with no
%% predefined interpretation of function symbols\footnote{A model
%%   specifies the universe and the interpretation of function symbols.},
%% was proved by a non-constructive model-theoretic argument
%% \cite{Lyndon59}. In the following, we shall consider only those
%% theories of the data domain $\Data$, for which there exists a decision
%% procedure for satisfiability, able to produce Lyndon interpolants. For
%% instance, such interpolating decision procedures exist for linear real
%% and integer arithmetic, with uninterpreted function symbols (UFLRA and
%% UFLIA).

@@

This is where the predicate-free path formulae (Definition
\ref{def:substform}) come into play. For a given event sequence
$\alpha$, the automaton $\A$ accepts a word $w$ such that $\event{w} =
\alpha$ if and only if $\substform{\alpha}$ is satisfiable. Assuming
further that the equality atoms in the transition rules of $\A$ are
written in the language of a decidable first order theory, such as
Presburger arithmetic, Lemma \ref{lemma:quant-pred-acceptance} gives
us an effective way of checking emptiness of $\A$, relative to a given
event sequence. However, this method does not cope well with lazy
annotation, because there is no way to extract, from the
unsatisfiability proof of $\substform{\alpha}$, the interpolants
needed to annotate $\alpha$. This is because \begin{inparaenum}[(i)]
\item\label{it1:gli} the formula $\substform{\alpha}$, obtained by
  repeated substitutions (Definition \ref{def:substform}) loses track
  of the steps of the execution, and
%
\item\label{it2:gli} quantifiers that occur nested in
  $\substform{\alpha}$ make it difficult to write $\substform{\alpha}$
  as an unsatisfiable conjunction of formulae from which interpolants
  are extracted (Definition \ref{def:lyndon-interpolant}).
\end{inparaenum}

@@

Under the assumption that witness terms can be effectively built, let
us describe the generation of a non-local GLI for a given input event
sequence $\alpha = a_1 \ldots a_n$. First, we generate successively
the acceptance formula $\accform{\alpha}$ and its equisatisfiable
forms $\quantform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \widehat{\Phi}$
and $\substform{\alpha} = Q_1x_1 \ldots Q_mx_m ~.~ \overline{\Phi}$,
both written in prenex form, with matrices $\widehat{\Phi}$ and
$\overline{\Phi}$, respectively. Because we assumed that the
first order theory of $\Data$ has quantifier elimination, the
satisfiability problem for $\substform{\alpha}$ is decidable. If
$\substform{\alpha}$ is satisfiable, we build a counterexample for
emptiness $w$ such that $\event{w}=\alpha$ and $\data{w}$ is a
satisfying assignment for $\substform{\alpha}$. Otherwise,
$\substform{\alpha}$ is unsatisfiable and there exist witness terms
$\tau_{i_1} \ldots \tau_{i_\ell}$, where $\set{i_1, \ldots, i_\ell} =
\set{j \in [1,m] \mid Q_j = \forall}$, such that
$\overline{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$ is unsatisfiable (Theorem
\ref{thm:skolem}). Then it turns out that the formula
$\widehat{\Phi}[\tau_{i_1}/x_{i_1}, \ldots,
  \tau_{i_\ell}/x_{i_\ell}]$, obtained analogously from the matrix of
$\quantform{\alpha}$, is unsatisfiable as well (Lemma
\ref{lemma:subst-quant}). Because this latter formula is structured as
a conjunction of formulae $\stamp{\iota}{0} \wedge \phi_1 \ldots
\wedge \phi_n \wedge \psi$, where $\voc{\phi_k} \cap \stamp{Q}{\leq n}
\subseteq \stamp{Q}{k-1} \cup \stamp{Q}{k}$ and $\voc{\psi} \cap
\stamp{Q}{\leq n} \subseteq \stamp{Q}{n}$, it is now possible to use
an existing interpolation procedure for the quantifier-free theory of
$\Data$, extended with uninterpreted function symbols, to compute a
sequence of non-local GLI $(I_0, \ldots, I_n)$ such that $\voc{I_k}
\cap \stamp{Q}{\leq n} \subseteq \stamp{Q}{k}$, for all $k \in [n]$.

@@

  we assumed that the first order theory of the data domain has

@@

In conclusion, under two assumptions about the first order theory of
the data domain, namely the \begin{inparaenum}[(i)]
\item witness-producing quantifier elimination, and
\item Lyndon interpolation for the quantifier-free fragment with
  uninterpreted functions,
\end{inparaenum}
we developped a rather generic method that produces generalized Lyndon
interpolants for unfeasible input event sequences. Moreover, each
formula $I_k$ in the interpolant refers only to the current predicate
symbols $\stamp{Q}{I_k}$, the current and past input variables
$\stamp{X}{\leq k}$ and the existentially quantified transition
variables introduced at the previous steps $\set{x_j \mid j <
  \xi_{\max}^{-1}(k), Q_j = \exists}$. The remaining question is how
to use such non-local interpolants to label the unfolding of an
automaton (Definition \ref{def:unfolding}) and to compute the coverage
between nodes of the unfolding (Definition \ref{def:coverage}).

@@

The main application of first order alternating automata is checking
inclusion between various classes of automata extended with variables
ranging over infinite domains that recognize languages over infinite
alphabets. The most widely known such classes are \emph{timed
  automata} \cite{AlurDill94} and \emph{finite-memory (register)
  automata} \cite{KaminskiFrancez94}. In both cases, complementation
is not possible inside the class and inclusion is undecidable. Our
contribution is providing a systematic semi-algorithm for these
decision problems. In addition, the method described in
\S\ref{sec:emptiness} can extend our previous \emph{generic register
  automata} \cite{IosifRV16} inclusion checking framework, by allowing
monitor (right-hand side) automata to have local variables, that are
not visible in the language.

Another application is checking safety (mutual exclusion, absence of
deadlocks, etc.) and liveness (termination, lack of starvation, etc.)
properties of parameterized concurrent programs, consisting of an
unbounded number of replicated threads that communicate via a fixed
set of global variables (locks, counters, etc.). The verification of
parametric programs has been reduced to checking the emptiness of a
(possibly infinite) sequence of first order alternating automata,
called \emph{predicate automata} \cite{Farzan15,Farzan16}, encoding
the inclusion of the set of traces of a parametric concurrent program
into increasingly general \emph{proof spaces}, obtained by
generalization of counterexamples. The program and the proof spaces
are first order alternating automata over the infinite alphabet of
pairs consisting of program statements and thread identifiers.
